\section{The time-dependent Hartree-Fock equations in a basis}
    We denote the basis of atomic orbitals (our original system) by
    $\brac{\ket{\chi_\alpha}}_{\alpha = 1}^L$. The Hartree-Fock wave function is
    thus described by the Slater determinant
    \begin{align}
        \ket{\Phi} = \brak{\phi_1, \phi_2, \dots, \phi_N},
    \end{align}
    where the basis of orthonormal molecular orbitals (Hartree-Fock
    spin-orbitals) is denoted by $\brac{\ket{\phi_i}}_{i = 1}^N$. They are
    represented as a linear combination of the atomic orbitals, viz.
    \begin{align}
        \ket{\phi_i} = C_{\alpha i}\ket{\chi_{\alpha}},
    \end{align}
    where we use the Einstein summation convention. In the time-dependent
    Hartree-Fock method we will evolve the molecular orbitals in time, but we
    keep the atomic orbitals fixed. This means that
    \begin{align}
        \ket{\phi_i(t)} = C_{\alpha i}(t)\ket{\chi_{\alpha}},
    \end{align}
    that is, the time-dependence occurs in the coefficients $C_{\alpha i}(t)$.
    Thus the time-dependent Hartree-Fock Slater determinant is given by
    \begin{align}
        \ket{\Phi(t)}
        &=
        \brak{
            \phi_1(t), \dots, \phi_N(t)
        }.
    \end{align}
    The time-dependent Slater determinants in a system with a time-dependent
    Hamiltonian $\hat{H}(t)$ satisfies the Schrödinger equation
    \begin{align}
        i\hslash\dpd[]{}{t}\ket{\Phi(t)}
        &= \hat{H}(t)\ket{\Phi(t)}.
    \end{align}
    The Hamiltonian is given by
    \begin{align}
        \hat{H}(t) &= \hat{h}(t) + \hat{u}(t),
    \end{align}
    where $\hat{h}(t)$ and $\hat{u}(t)$ is the time-dependent one-body and
    two-body operators respectively.

    \subsection{Deriving the time-dependent Hartree-Fock equations}
        For the sake of brevity we will remove the explicit time-dependence in
        the orbitals, the Slater determinants and the operators, though they are
        still present. We will be following closely the derivation done by
        \citeauthor{hochstuhl2014time}, but for a single Slater determinant. The
        expectation value of the Hamiltonian in terms of the Slater determinant
        is then
        \begin{align}
            \bra{\Phi}\hat{H}\ket{\Phi}
            &= \bra{\phi_i}\hat{h}\ket{\phi_i}
            + \frac{1}{2}\bra{\phi_i\phi_j}\hat{u}\ket{\phi_i\phi_j}_{AS},
        \end{align}
        where the antisymmetric matrix elements are given by
        \begin{align}
            \bra{\phi_i\phi_j}\hat{u}\ket{\phi_i\phi_j}_{AS}
            &= \bra{\phi_i\phi_j}\hat{u}\ket{\phi_i\phi_j}
            - \bra{\phi_i\phi_j}\hat{u}\ket{\phi_j\phi_i}.
            \label{eq:antisymmetric_two_body}
        \end{align}
        Furthermore, the time derivative of the Slater determinant is
        % TODO: How do you get this result?
        % Write out SD as sum, and use orthonormality of orbitals?
        % Seems like only one term "survives"
        \begin{align}
            \bra{\Phi}\dpd{}{t}\ket{\Phi}
            &= \bra{\phi_i}\dpd{}{t}\ket{\phi_i}.
        \end{align}
        Introducing functional derivatives, which when acting on the matrix
        elements gives
        \begin{gather}
            \dfd{\phi_k^{*}}\bra{\phi_i}\hat{h}\ket{\phi_i}
            = \hat{h}\ket{\phi_k}, \\
            \dfd{\phi_k^{*}}\bra{\phi_i}\dpd{}{t}\ket{\phi_i}
            = \dpd{}{t}\ket{\phi_i}, \\
            \dfd{\phi_k^{*}}\bra{\phi_i\phi_j}\hat{u}\ket{\phi_i\phi_j}_{AS}
            = 2\bra{\cdot\phi_j}\hat{u}\ket{\phi_k\phi_j}_{AS},
        \end{gather}
        where the "dot" is defined in terms of the tensor product by
        \begin{align}
            \ket{\phi_i\phi_j} = \ket{\phi_i}\otimes\ket{\phi_j}
            \implies \ket{\cdot\phi_j} \equiv \1 \otimes\ket{\phi_j}.
        \end{align}
        The last derivative can be found by expanding the antisymmetrix matrix
        elements by \autoref{eq:antisymmetric_two_body} and applying the
        functional derivative two times on each term. Collecting the four terms
        and using the symmetric property of the two-body elements that
        \begin{align}
            \bra{\phi_i\phi_j}\hat{u}\ket{\phi_k\phi_l}
            = \bra{\phi_j\phi_i}\hat{u}\ket{\phi_l\phi_k}.
        \end{align}
        We can now find the equations of motion by applying time-dependent
        variational principle, that is,
        \begin{align}
            \bra{\delta\Phi}\hat{H} - i\hslash\dpd{}{t}\ket{\Phi} = 0,
        \end{align}
        under the constraint that the time-dependent molecular orbitals are
        orthonormal, i.e.,
        \begin{align}
            \braket{\phi_i(t)}{\phi_j(t)} = \delta_{ij},
        \end{align}
        where we for the sake of emphasis temporarily added the explicit
        time-dependence in the orbitals. We now define the Lagrangian functional
        with the Lagrange multipliers $\lambda_{ij}$ to be
        \begin{align}
            \mathcal{L}
            &= \bra{\Phi}\hat{H} - i\hslash\dpd{}{t}\ket{\Phi}
            - \lambda_{ij}\para{
                \braket{\phi_i}{\phi_j} - \delta_{ij}
            }.
        \end{align}
        We now find an extremum of $\mathcal{L}$ by taking the functional
        derivative of the functional and equating it to zero.
        \begin{gather}
            \dfd{\phi_k^{*}}\mathcal{L} = 0, \\
            \implies
            \hat{h}\ket{\phi_k}
            + \bra{\cdot\phi_j}\hat{u}\ket{\phi_k\phi_j}_{AS}
            - i\hslash\dpd{}{t}\ket{\phi_k}
            - \lambda_{kj}\ket{\phi_j} = 0.
            \label{eq:extremum_lagrangian}
        \end{gather}
        Left-projecting with an orbital $\bra{\phi_l}$ and solving for the
        Lagrange multipliers, $\lambda_{kl}$, where we apply the constraint that
        the orbitals are orthonormal we get
        \begin{align}
            \lambda_{kl}
            =
            \bra{\phi_l}\hat{h}\ket{\phi_k}
            + \bra{\phi_l\phi_j}\hat{u}\ket{\phi_k\phi_j}_{AS}
            - i\hslash\bra{\phi_l}\dpd{}{t}\ket{\phi_k}.
            \label{eq:lagrange_multipliers}
        \end{align}
        Inserting the expression for the Lagrange multipliers in
        \autoref{eq:lagrange_multipliers} into \autoref{eq:extremum_lagrangian}
        we get
        \begin{align}
            \hat{P}\para{
                \hat{h}\ket{\phi_k}
                + \bra{\cdot\phi_j}\hat{u}\ket{\phi_k\phi_j}_{AS}
                - i\hslash\dpd{}{t}\ket{\phi_k}
            } = 0,
        \end{align}
        where we have defined the projection operator, $\hat{P}$, to be
        \begin{align}
            \hat{P} = \1 - \ket{\phi_i}\bra{\phi_i}.
        \end{align}
        Solving for the time-derivative of the molecular orbitals we get
        \begin{align}
            i\hslash\hat{P}\dpd{}{t}\ket{\phi_k}
            = \hat{P}\para{
                \hat{h}\ket{\phi_k}
                + \bra{\cdot\phi_j}\hat{u}\ket{\phi_k\phi_j}_{AS}
            }
            = \hat{P}\hat{f}\ket{\phi_k},
            \label{eq:tdhf_equations_p}
        \end{align}
        where we have defined the Fock operator to be
        \begin{align}
            \hat{f}
            \equiv \hat{h} + \bra{\cdot\phi_j}\hat{u}\ket{\cdot\phi_j}_{AS}.
        \end{align}
        We now define an arbitrary hermitian operator $\hat{Q}(t)$ in terms of
        a unitary transformation
        \begin{align}
            \bra{\phi_i}i\hslash\dpd{}{t}\ket{\phi_j}
            \equiv \bra{\phi_i}\hat{Q}(t)\ket{\phi_j}.
            \label{eq:unitary_transformation_time}
        \end{align}
        This is valid as any Slater determinant is invariant under unitary
        transformations. See section \ref{ssub:slater_determinants_invariant}
        for a short proof of this. As the Slater determinant is invariant under
        the unitary transformation, by extension the Lagrangian functional will
        be invariant. Expanding $\hat{P}$ in
        \autoref{eq:tdhf_equations_p} and inserting
        \autoref{eq:unitary_transformation_time} we get
        \begin{gather}
            i\hslash\dpd{}{t}\ket{\phi_k}
            - i\hslash\ket{\phi_i}\bra{\phi_i}\dpd{}{t}\ket{\phi_k}
            = \hat{f}\ket{\phi_k}
            - \ket{\phi_i}\bra{\phi_i}\hat{f}\ket{\phi_k}
            \\
            \implies
            i\hslash\dpd{}{t}\ket{\phi_k}
            - \ket{\phi_i}\bra{\phi_i}\hat{Q}(t)\ket{\phi_k}
            = \hat{f}\ket{\phi_k}
            - \ket{\phi_i}\bra{\phi_i}\hat{f}\ket{\phi_k}.
        \end{gather}
        Since $\hat{Q}(t)$ was an arbitrary hermitian operator and
        $\hat{f}^{\dagger}(t) = \hat{f}(t)$ we can choose $\hat{Q}(t) =
        \hat{f}(t)$ yielding the time-dependent Hartree-Fock equations on the
        form
        \begin{align}
            i\hslash\dpd{}{t}\ket{\phi_k}
            = \hat{f}\ket{\phi_k}.
        \end{align}
        We will in the next section restore the explicit time-dependence in the
        molecular orbitals and the operators.

    \subsection{Evolving the coefficients in time}
        Having restricted the full wave function to a single Slater determinant,
        we can write the equations for the molecular orbitals in terms of the
        time-dependent Fock operator.
        \begin{align}
            i\hslash\dpd{}{t}\ket{\phi_i(t)}
            &= \hat{f}(t)\ket{\phi_i(t)}.
        \end{align}
        We can now insert the linear combination for the molecular orbitals in
        the atomic orbital basis with the time-dependent coefficients.
        \begin{align}
            i\hslash\dpd{}{t}C_{\alpha i}(t)\ket{\chi_{\alpha}}
            &= \hat{f}(t)C_{\alpha i}(t)\ket{\chi_{\alpha}}.
        \end{align}
        Left projecting with another state from the atomic orbitals we can rewrite
        the previous equation to
        \begin{gather}
            i\hslash\dpd{}{t}C_{\alpha i}(t)\braket{\chi_{\beta}}{\chi_{\alpha}}
            =
            C_{\alpha i}(t)\bra{\chi_{\beta}}\hat{f}(t)\ket{\chi_{\alpha}}
            \\
            \implies
            i\hslash \dot{C}_{\alpha i} S_{\beta\alpha}
            = C_{\alpha i}(t)f_{\beta\alpha}(t)
            \\
            \implies
            i\hslash \vf{S}\dot{\vf{C}}
            = \vf{F}(t)\vf{C}(t),
            \label{eq:tdhf-equations}
        \end{gather}
        where the overlap matrix $\vf{S}$ is the matrix with elements
        \begin{align}
            S_{\beta\alpha} \equiv \braket{\chi_{\beta}}{\chi_{\alpha}},
            \label{eq:overlap_ao}
        \end{align}
        and the time-dependent Fock matrix $\vf{F}(t)$ contains the entries
        \begin{align}
            f_{\beta\alpha}(t) = \bra{\chi_{\beta}}\hat{f}(t)\ket{\chi_\alpha}.
        \end{align}
        Often we will work with basis sets where the overlap matrix reduces to
        the identity. In this case, \autoref{eq:tdhf-equations} collapses to the
        slightly simpler equation
        \begin{align}
            i\hslash \dot{\vf{C}} = \vf{F}(t)\vf{C}(t).
            \label{eq:tdhf-orthogonal}
        \end{align}


        \subsubsection{Proof that Slater determinants are invariant under
        unitary transformations}
            \label{ssub:slater_determinants_invariant}
            Given a basis $\brac{\ket{\phi_i}}_{i = 1}^{N}$ that spans the
            $N$-dimensional Hilbert space. We can then do a unitary
            transformation from this basis to a new basis
            $\brac{\ket{\psi_i}}_{i = 1}^{N}$ in the same Hilbert space by
            \begin{align}
                \ket{\psi_i} = u_{ji}\ket{\phi_j},
                \label{eq:unitary_transformation}
            \end{align}
            where $u_{ji} = u_{ij}^{*}$. For a set of coordinates
            $\brac{\vf{x}_1, \dots, \vf{x}_N}$, we write
            \begin{align}
                \psi_{ij} \equiv \psi_j(\vf{x}_i)
                \equiv \braket{\vf{x}_i}{\psi_j},
            \end{align}
            and equivalently for $\phi_{ij}$. Projecting onto the coordinate
            basis we can write \autoref{eq:unitary_transformation} as
            \begin{gather}
                \braket{\vf{x}_k}{\psi_i}
                = u_{ji}\braket{\vf{x}_k}{\phi_j}
                \implies
                \psi_{ki} = u_{ji}\phi_{kj} = \phi_{kj}u_{ji}
                \implies \vfg{\Psi} = \vfg{\Phi}\vfg{U},
            \end{gather}
            where the matrices $\vfg{\Psi}$, $\vfg{\Phi}$ and $\vfg{U}$ are
            the matrices with elements
            \begin{align}
                \vfg{\Psi}
                &= \begin{pmatrix}
                    \psi_{11} & \psi_{12} & \dots & \psi_{1N} \\
                    \vdots & \vdots & \ddots & \vdots \\
                    \psi_{N1} & \psi_{N2} & \dots & \psi_{NN}
                \end{pmatrix}, \\
                \vfg{\Phi}
                &= \begin{pmatrix}
                    \phi_{11} & \phi_{12} & \dots & \phi_{1N} \\
                    \vdots & \vdots & \ddots & \vdots \\
                    \phi_{N1} & \phi_{N2} & \dots & \phi_{NN}
                \end{pmatrix}, \\
                \vfg{U}
                &= \begin{pmatrix}
                    u_{11} & u_{12} & \dots & u_{1N} \\
                    \vdots & \vdots & \ddots & \vdots \\
                    u_{N1} & u_{N2} & \dots & u_{NN}
                \end{pmatrix},
            \end{align}
            and where $U^{\dagger}U = \1$. Creating the fully antisymmetrized
            normalized wave function, i.e., the Slater determinant, from
            $\vfg{\Psi}$ and $\vfg{\Phi}$ we see that
            \begin{gather}
                \det(\vfg{\Psi}) = \det(\vfg{\Phi}\vfg{U})
                = \det(\vfg{\Phi})\det(\vfg{U})
            \end{gather}
            We now take the squared norm (any vector norm) on both sides.
            \begin{align}
                \norm{\det(\vfg{\Psi})}^2
                = \norm{\det(\vfg{\Phi})\det(\vfg{U})}^2
                \leq
                \norm{\det(\vfg{\Phi})}^2\norm{\det(\vfg{U})}^2
                = \norm{\det(\vfg{\Phi})}^2,
                \label{eq:squared_determinant}
            \end{align}
            where we have used that
            \begin{align}
                \norm{\det(\vfg{U})}^2
                &= \norm{\det(\vfg{U^{\dagger}})\det(\vfg{U})}
                = \norm{\det(\vfg{U}^{\dagger}\vfg{U})}
                = \norm{\det(\1)} = 1.
            \end{align}
            Now, in \autoref{eq:squared_determinant}, since the Slater
            determinant of $\vfg{\Phi}$ is orthonormalized and a unitary
            transformation preserves the normalization this means that the norm
            of $\det(\vfg{\Phi})$ and the norm of $\det(\vfg{\Psi})$ must be
            unity.  Thus, the equality in the splitting of the norm is preserved
            thus yielding
            \begin{gather}
                \norm{\det(\vfg{\Psi})}^2 = \norm{\det(\vfg{\Phi})}^2.
            \end{gather}



    \subsection{Time evolution in a non-orthogonal basis of atomic orbitals}
        In order to solve \autoref{eq:tdhf-orthogonal} for non-orthogonal
        atomic orbitals and avoid having to solve \autoref{eq:tdhf-equations}
        as they stand, we orthogonalize the atomic orbitals. We will follow the
        derivation shown in the book \citetitle{szabo1996modern} by
        \citeauthor{szabo1996modern} \cite{szabo1996modern} closely.

        For the case when
        \begin{align}
            \braket{\chi_{\alpha}}{\chi_{\beta}} \equiv S_{\alpha\beta}
            \neq \delta_{\alpha\beta},
        \end{align}
        that is, a non-orthogonal atomic orbital basis
        $\brac{\ket{\chi_{\alpha}}}_{\alpha = 1}^{N}$, we move to a transformed
        orthonormal basis $\brac{\ket{\xi_{\mu}}}_{\mu = 1}^{K}$ by the
        transformation
        \begin{align}
            \ket{\xi_{\mu}} = X_{\alpha\mu}\ket{\chi_{\alpha}}.
        \end{align}
        This yields an equation on the form
        \begin{align}
            \braket{\xi_{\mu}}{\xi_{\nu}}
            &= \para{
                X_{\alpha\mu}^{*}\bra{\chi_{\alpha}}
            }\para{
                X_{\beta\nu}\ket{\chi_{\beta}}
            }
            = X_{\alpha\mu}^{*}S_{\alpha\beta}X_{\beta\nu}
            = \delta_{\mu\nu}.
        \end{align}
        As a matrix equation we get
        \begin{align}
            \vf{X}^{\dagger}\vf{S}\vf{X} = \1.
            \label{eq:orthonormal-transformation}
        \end{align}
        This, along with the requirement that $\vf{X}$ must be non-singular,
        forms the restrictions on the matrix $\vf{X}$ in order for
        $\brac{\ket{\xi_{\mu}}}_{\mu = 1}^{L}$ to form an orthonormal basis.
        Since $\vf{S}$ is Hermitian (by construction), there exists a unitary
        matrix $\vf{U}$ that diagonalizes $\vf{S}$ by
        \begin{align}
            \vf{U}^{\dagger}\vf{S}\vf{U} = \vf{s},
        \end{align}
        where $\vf{s}$ is a diagonal matrix with the eigenvalues of $\vf{S}$.
        The matrix $\vf{S}$ is positive definite as all the eigenvalues of
        $\vf{S}$ are positive. From \citeauthor{mayer2002lowdin} we can find an
        equation for $\vf{X}$ using \emph{symmetric orthogonalization}. This
        yields
        \begin{align}
            \vf{X} = \vf{S}^{-1/2} = \vf{U}\vf{s}^{-1/2}\vf{U}^{\dagger}.
        \end{align}
        We can now use $\vf{X}$ to transform to the new basis
        $\brac{\xi_{\mu}}_{\mu = 1}^{L}$ and solving the Roothan-Hall equations
        in this orthonormal basis. In practice we do this by transforming the
        coefficient matrix $\vf{C}$ from the atomic orbitals in terms of the
        orthonormal orbitals $\xi_{\mu}$. In terms of the original atomic
        orbitals we can formulate this by
        \begin{align}
            \ket{\phi_i}
            &= C_{\mu i}'\ket{\xi_{\mu}}
            = C_{\mu i}'X_{\alpha \mu}\ket{\chi_{\alpha}}
            = C_{\alpha i}\ket{\chi_{\alpha}},
        \end{align}
        where $C_{\mu i}'$ are the coefficients from the new orthonormal orbital
        basis $\ket{\xi_{\mu}}$ to the molecular orbitals. In terms of the
        coefficient matrices we then get
        \begin{align}
            \vf{C} = \vf{X}\vf{C}',
        \end{align}
        where $\vf{C}$ are the original coefficients from the original atomic
        orbital basis $\ket{\chi_{\alpha}}$ to the molecular orbital basis.
        Having eliminated the linear dependencies in $\vf{S}$ there will exist
        an inverse for $\vf{X}$. Thus we find
        \begin{align}
            \vf{C}' = \vf{X}^{-1}\vf{C}.
        \end{align}
        Inserting this expression into \autoref{eq:tdhf-equations} when the
        coefficient matrices are time-dependent\footnote{Note that $\vf{X}(t) =
        \vf{X}$ as we keep the atomic orbitals fixed and let the coefficients
        handle the time-dependence.} we get
        \begin{gather}
            i\hslash\vf{S}\dot{\vf{C}}(t) = \vf{F}(t)\vf{C}(t) \\
            \implies
            i\hslash\vf{S}\vf{X}\dot{\vf{C}}'(t) = \vf{F}(t)\vf{X}\vf{C}'(t)
            \\
            \implies
            i\hslash\vf{X}^{\dagger}\vf{S}\vf{X}\dot{\vf{C}}'(t)
            = \vf{X}^{\dagger}\vf{F}(t)\vf{X}\vf{C}'(t)
            \\
            \implies
            i\hslash \dot{\vf{C}}'(t) = \vf{F}'(t)\vf{C}'(t),
        \end{gather}
        where we've used the orthonormality of the transformation in
        \autoref{eq:orthonormal-transformation} and defined the transformed Fock
        matrix
        \begin{align}
            \vf{X}^{\dagger}\vf{F}(t)\vf{X} = \vf{F}'(t).
        \end{align}
        Doing this transformation simplifies the use of linear algebra libraries
        to solve the time-dependent Hartree Fock equations.

    \subsection{Time dependent overlap}
        When working with Slater determinants in a given basis we can compute
        the overlap between two wavefunctions in an efficient manner. We define
        the orbitals
        \begin{align}
            \ket{\phi_i} &= C_{\alpha i}\ket{\chi_{\alpha}}, \\
            \ket{\psi_i} &= D_{\alpha i}\ket{\chi_{\alpha}},
        \end{align}
        as two separate linear combinations of the original atomic orbitals. The
        overlap between these two molecular orbitals is then
        \begin{align}
            \braket{\psi_i}{\phi_j}
            &= \para{
                D_{\alpha i}^{*}
                \bra{\chi_{\alpha}}
            }\para{
                C_{\beta j}
                \ket{\chi_{\beta}}
            }
            = D_{\alpha i}^{*}S_{\alpha\beta} C_{\beta j}
            = T_{ij}.
        \end{align}
        In matrix notation this reduces to
        \begin{align}
            \vf{T} = \vf{D}^{\dagger}\vf{S}\vf{C},
        \end{align}
        where $\vf{C}$ and $\vf{D}$ are the coefficient matrices of the set of
        molecular orbitals $\ket{\phi_i}$ and $\ket{\psi_i}$ respectively and
        $\vf{S}$ is the overlap matrix for the atomic orbitals.  We can then
        write the two Slater determinants consisting of $N$ of the molecular
        orbitals as
        \begin{align}
            \ket{\Phi} &= \frac{1}{\sqrt{N!}}
            \sum_{\sigma \in S_N}(-1)^{|\sigma|}
            \ket{\phi_{\sigma(1)}}
            \otimes\dots\otimes
            \ket{\phi_{\sigma(N)}}, \\
            \ket{\Psi} &= \frac{1}{\sqrt{N!}}
            \sum_{\sigma \in S_N}(-1)^{|\sigma|}
            \ket{\psi_{\sigma(1)}}
            \otimes\dots\otimes
            \ket{\psi_{\sigma(N)}},
        \end{align}
        where $S_N$ is the permutation group of all permutations $\sigma$ of the
        labels $i \in \brac{1, \dots, N}$. Furthermore, $|\sigma|$, is the
        defined as the number of permutations performed from the original order
        of the labels. The overlap of the two Slater determinants is then
        \begin{align}
            \braket{\Psi}{\Phi}
            &= \frac{1}{N!}
            \sum_{\sigma, \tau \in S_{N}}
            (-1)^{|\sigma||\tau|}
            \prod_{i = 1}^{N}
            \braket{\psi_{\sigma(i)}}{\phi_{\tau(i)}}
            \\
            &= \frac{1}{N!}
            \sum_{\sigma, \tau \in S_{N}}
            (-1)^{|\sigma||\tau|}
            \prod_{i = 1}^{N}
            T_{\sigma(i)\tau(i)},
        \end{align}
        where we now have a double sum over all permutations $\sigma$ and
        $\tau$. To go from here we have to collapse one of the sums over the
        permutations. By keeping one of the permutation sums fixed we get all
        unique combinations of the terms in the products. By iterating over the
        second sum over permutations we create the same set of combinations, but
        with a different ordering. As the terms commute, we can just add it to
        the already existing set of unique combinations. In total there are $N!$
        permutations in $S_N$ thus giving a factor of $N!$ to the sum. This
        yields
        \begin{align}
            \braket{\Psi}{\Phi}
            &= \frac{1}{N!}N!
            \sum_{\sigma \in S_N}
            (-1)^{|\sigma|}
            \prod_{i = 1}^{N}T_{1\sigma(1)}
            \\
            &= \begin{vmatrix}
                T_{11} & \dots & T_{1N} \\
                \vdots & \ddots & \vdots \\
                T_{N1} & \dots & T_{NN}
            \end{vmatrix}
            = \det(\vf{T})
        \end{align}
        In particular in the time-dependent Hartree-Fock regime where we have
        the time-dependent coefficients $C_{\alpha i}(t)$ we get the overlap
        with the ground state to be
        \begin{align}
            \braket{\Phi(t)}{\Phi(0)}
            &= \det(\vf{C}^{\dagger}(t)\vf{S}\vf{C}(0))
            = \det(\vf{T}(t)),
        \end{align}
        where $\vf{C}(t)$ is the coefficient matrix of the time evolved states.

        \subsubsection{Overlap with non-orthogonal atomic orbitals}
            In the case that we have transformed from the non-orthogonal atomic
            orbital basis $\brac{\ket{\chi_{\alpha}}}_{\alpha = 1}^{N}$ to the
            orthonormal atomic orbital basis $\brac{\ket{\xi_{\mu}}}_{\mu =
            1}^{N}$ by,
            \begin{align}
                \ket{\xi_{\mu}} = X_{\alpha\mu}\ket{\chi_{\alpha}},
            \end{align}
            where the transformation to the molecular orbitals is given by,
            \begin{align}
                \ket{\phi_{i}(t)} = C_{\mu i}'(t)\ket{\xi_{\mu}}
                = C_{\mu i}'(t)X_{\alpha\mu}\ket{\chi_{\alpha}}
                = C_{\alpha i}(t)\ket{\chi_{\alpha}},
            \end{align}
            we get the time-dependent overlap to be
            \begin{align}
                \braket{\Phi(t)}{\Phi(0)}
                &= \det(\vf{C}^{\dagger}(t)\vf{S}\vf{C}(0))
                \\
                &= \det\Bigl[
                    (\vf{C}'(t))^{\dagger} \vf{X}^{\dagger}
                    \vf{S}\vf{X}\vf{C}'(t)
                \Bigr]
                \\
                &=
                \det\Bigl[
                    (\vf{C}'(t))^{\dagger}
                    \vf{C}'(t)
                \Bigr].
            \end{align}
            This simplifies the process of computing the overlap, when we've
            already transformed to the orthonormal basis.

    \subsection{The time evolution operators}
        We will in the following subsection stay close to the derivation of
        \citeauthor{ullrich2011time}, section 3.1.2, and
        \citeauthor{joachain2012atoms}, section 5.1.  Any solution to the
        time-dependent Schrödinger equation is given by
        \begin{align}
            \ket{\Psi(t)} = \hat{U}(t, t_0)\ket{\Psi(t_0)},
        \end{align}
        where $\hat{U}(t, t_0)$ is the time evolution operator that acts on the
        initial state $\ket{\Psi(t_0)}$, and yields the state $\ket{\Psi(t)}$ at
        some later time $t$. The time evolution operators are unitary at common
        times, viz.
        \begin{align}
            \hat{U}^{\dagger}(t, t_0)\hat{U}(t, t_0) = \1.
        \end{align}
        This means that we can go ``backwards in time'' in the sense that going
        from $\ket{\Psi(t)} \to \ket{\Psi(t_0)}$ is done by
        \begin{align}
            \ket{\Psi(t_0)} = \hat{U}^{\dagger}(t, t_0)\ket{\Psi(t)}.
        \end{align}
        Furthermore, we can compose a time evolution operator from other time
        evolution operators as
        \begin{align}
            \hat{U}(t_2, t_0) = \hat{U}(t_2, t_1)\hat{U}(t_1, t_0),
        \end{align}
        where $t_2 \geq t_1 \geq t_0$. If the Hamiltonian is time-independent,
        i.e., $\hat{H}(t) = \hat{H}$, the time evolution operator takes on the
        form
        \begin{align}
            \hat{U}(t, t_0) = \exp\brac{
                \frac{-i \hat{H} (t - t_0)}{\hslash}
            }.
            \label{eq:ti-evolution}
        \end{align}
        However, if the Hamiltonian is time-dependent, the time evolution
        operator takes on a much more complicated shape.
        \begin{align}
            \hat{U}(t, t_0) =
            \mathcal{T}\exp\brac{
                -\frac{i}{\hslash} \int_{t_0}^{t} \dd\tau
                \hat{H}(\tau)
            },
            \label{eq:td-evolution}
        \end{align}
        where $\mathcal{T}$ is the time-ordering operator. As an extra
        complicating factor, the time-dependent Hamiltonian might not commute
        with itself at different times. Here \autoref{eq:ti-evolution} and
        \autoref{eq:td-evolution} serve as the theoretical foundation for how
        the time-evolution is practically done in a numerical scheme.

    \subsection{Numerical time propagation using the Crank-Nicolson algorithm}
        Continuing the derivation from section 4.5 in
        \citeauthor{ullrich2011time}, we look at how we can numerically
        propagate in time a wave function with a time-dependent Hamiltonian.
        Luckily, due to the property of the time evolution operators that they
        can be composed of several intermediate time steps, discretizing the
        operators becomes quite natural. That is, if we define $\tau_1 = t_0$,
        and $\tau_N = t$ with all intermediate times given by
        \begin{align}
            \tau_{j + 1} = \tau_j + \Delta \tau,
        \end{align}
        where the time step $\Delta \tau$ is given by
        \begin{align}
            \Delta \tau = \frac{t - t_0}{N - 1},
        \end{align}
        and $N$ is the number of discrete time steps. This means that we can
        write
        \begin{align}
            \ket{\Psi(\tau_{j + 1})}
            = \hat{U}(\tau_{j + 1}, \tau_{j})\ket{\Psi(\tau_j)},
        \end{align}
        where the equality is maintained. For a time-dependent Hamiltonian
        (which is what we will be working on), the time evolution operator is
        given by \autoref{eq:td-evolution}, but this is intractable in a
        numerical scheme, to say the least. We therefore approximate the
        operator by
        \begin{align}
            \hat{U}(\tau_{j + 1}, \tau_{j})
            \approx \exp\brac{
                -\frac{i}{\hslash}\hat{H}(\tau_{j} + \Delta\tau/2)\Delta\tau
            }
            \equiv
            \exp\brac{
                -\frac{i}{\hslash}\hat{H}(\tau_{j + 1/2})\Delta\tau
            },
        \end{align}
        that is, we evaluate the Hamiltonian at the midpoint between $\tau_{j +
        1}$ and $\tau_{j}$. The choice of evaluating the time step at the
        midpoint comes about in order to preserve the unitarity of the
        approximated time evolution operator. That is,
        \begin{align}
            \ket{\Psi(\tau_j)}
            = \hat{U}(\tau_j, \tau_{j + 1})\ket{\Psi(\tau_{j + 1})}
            = \hat{U}(\tau_j, \tau_{j + 1})\hat{U}(\tau_{j + 1}, \tau_j)
            \ket{\Psi(\tau_j)},
        \end{align}
        which means that
        \begin{align}
            \hat{U}(\tau_j, \tau_{j + 1})
            \hat{U}(\tau_{j + 1}, \tau_j)
            = \hat{U}^{\dagger}(\tau_{j + 1}, \tau_j)
            \hat{U}(\tau_{j + 1}, \tau_j)
            = \1.
        \end{align}
        Using the Crank-Nicolson algorithm, we can approximate the exponential
        in the approximated time evolution operator to be
        \begin{align}
            \hat{U}(\tau_{j + 1}, \tau_j)
            \approx \exp\brac{
                -\frac{i}{\hslash}\hat{H}(\tau_{j + 1/2})\Delta \tau
            }
            \approx
            \frac{1 - i\hat{H}(\tau_{j + 1/2})\Delta\tau/(2\hslash)}
            {1 + i\hat{H}(\tau_{j + 1/2})\Delta\tau/(2\hslash)},
        \end{align}
        which is correct to second order in $\Delta \tau$ and preserves
        unitarity. Expressed as system of linear equations, we then have
        \begin{align}
            \para{
                1 + \frac{i}{2\hslash}\hat{H}(\tau_{j + 1/2})\Delta\tau
            }\ket{\Psi(\tau_{j + 1})}
            =
            \para{
                1 - \frac{i}{2\hslash}\hat{H}(\tau_{j + 1/2})\Delta\tau
            }\ket{\Psi(\tau_{j})}.
        \end{align}
        We can either solve this equation by inversion, or as a system of linear
        equations. The latter will be more stable.
