This thesis would not have come about where it not for my two excellent
supervisors: Håkon Emil Kristiansen and Morten Hjorth-Jensen.
I would like to thank Håkon for allowing me to continue on his work, and for
including me in his own research.
Your open-mindedness towards new suggestions on any topic is truly mind-boggling
and inspiring for someone as stubborn as myself.
I am indebted to your counsel and friendship throughout this work.
I would also like to thank Morten for providing me with such an interesting
topic, and for all the guidance you have given me throughout the work with this
thesis.
Your insight, fascination, and knowledge on all things related to many-body
physics and computations is awe-inspiring.
Not to mention your passion for teaching and the well-being of students at the
Computational Science group.
I look forward to continued collaboration with both of you in the future.

Working with Sebastian Gregorius Winther-Larsen on this topic has made the work
so much more fun and productive.
I believe that we together have gotten so much farther than what would have been
possible if we went our separate ways.
I have truly enjoyed these last two years, both as partners in crime on our
theses, but perhaps mostly due to all the shenanigans.
Our trips to the US and to the UK were a blast, and I hope that we get to
continue all our ramblings in the next four years as PhD candidates.

So many hours have been spent at the old Computational Physics group, and most
of them were with my long-time office partners Alocias, Magnus, and Vilde.
Thank you all for an exciting and memorable time.
A huge thanks goes out to all the people at the Computational Science group for
making my time here so pleasant.
I would also like to thank the guys and gals at OSI Elvepadling for bringing me
along on adventures in the rivers all over Norway.
I am really itching to go paddling with you guys again.

To my brother, Vemund, who I always enjoy spending time with!
To my parents, Helle and Sigmund, for always being there and always rooting for
Vemund and me.
To my girlfriend, Jenny, for your support and patience throughout these last
months.
I feel so lucky to have you in my life, and I cannot wait to spend my future
with you.
I love you!

Finally, to the reader of this text, wherein there are no statements of topics
or derivations being ``easy'', ``simple'', or ``trivial''.
If a topic takes you several years of advanced physics study to understand, then
it is \emph{emphatically not} easy \cite{nontrivial-manifesto}.
